---
title: "On using the R package `tmle.npvi`"
author: "Nima Hejazi"
date: "`r Sys.Date()`"
output: html_document
---

```{r get-pkg, eval=FALSE}
library(devtools)
remove.packages("tmle.npvi")
install_github("achambaz/tmle.npvi")
```

Now, let's simply generate some data to be used with the `tmle.npvi` package.
The observed data structure will be of the form $O = (Y, X, W)$, with each of
the variables of interest being of the form: $Y \sim N(0, 1)$, $X \sim N(0, 1)$,
and $W = (W_1, W_2, W_3)$, with $W_i \sim N(0, 1)$. In order to ensure that the
exposure/treatment variable of interest ($X$) has sufficient mass at a null
value, we set $X \in [-1, 1] \mapsto 0$: that is, all observed values of $X$
within _one standard deviation_ of the mean are set to _zero_.

```{r sim}
suppressMessages(library(tmle.npvi))
set.seed(46781)
n_obs <- 1000
n_var <- 5

obs <- as.data.frame(replicate(n_var, rnorm(n_obs)))
colnames(obs) <- c("Y", "X", paste0("W", seq_len(n_var - 2)))
obs$X[obs$X > -1 & obs$X < 1] <- 0
head(obs)
```

Above, we examine the `data.frame` that represents the observed data structure
and confirm that it appears to meet the specifications required for invoking the
`tmle.npvi` procedure.

Now, we use the eponymous function of the `tmle.npvi` package to obtain a
targeted minimum loss-based estimate of the NPVI parameter. We then
appropriately set the confidence level and examine the output structure.

```{r npvi}
alpha <- 0.05
npvi <- tmle.npvi(obs, f = identity, flavor = "learning")
setConfLevel(npvi, 1 - alpha)
npvi
```

The object of class `npvi` returned by the estimation procedure provides a
wealth of information. We are interested in the point estimate of the estimator
of the parameter of interest $\psi$ as well as the uncertainty (as measured by
standard deviation) associated with this estimate.

__This leads to a few questions:__

1. While the point estimate is rather small ($-0.0621$) the standard error of
   the estimate is roughly an order of magnitude larger. Why is this?
2. While we note that the standard deviation of the estimator of $\psi$ is
   readily accessible via the `tmle.npvi::getPsiSd` function, this value is not
   used in computing the confidence interval (rather the standard error is) --
   why is this?
3. From this simulated data, we note that there is not a straightforward
   relationship between the standard deviation and standard error, contrary to
   our expectations. Why is this the case?

In particular, we would expect the confidence interval for the estimator of
$\psi$ to be computable as follows:

```{r npvi-stats}
psi <- getPsi(npvi)
sd <- getPsiSd(npvi)
CI = psi + c(-1,1) * sd * qnorm(1 - alpha/2) / sqrt(n_obs)
out <- c(CI[1], psi, CI[2])
names(out) <- c("CI_lower", "Est", "CI_upper")
out
```

Examining this CI, we note that it is considerably tighter than that reported by
the `print` method associated with the `npvi` object we computed.

