---
output: github_document
bibliography: refs.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# R/`methyvim`

[![Travis-CI Build Status](https://travis-ci.org/nhejazi/methyvim.svg?branch=master)](https://travis-ci.org/nhejazi/methyvim)
[![AppVeyor Build  Status](https://ci.appveyor.com/api/projects/status/github/nhejazi/methyvim?branch=master&svg=true)](https://ci.appveyor.com/project/nhejazi/methyvim)
[![Coverage Status](https://img.shields.io/codecov/c/github/nhejazi/methyvim/master.svg)](https://codecov.io/github/nhejazi/methyvim?branch=master)
[![Project Status: WIP - Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](http://www.repostatus.org/badges/latest/wip.svg)](http://www.repostatus.org/#wip)
[![MIT license](http://img.shields.io/badge/license-MIT-brightgreen.svg)](http://opensource.org/licenses/MIT)

> Targeted learning of variable importance measures for the analysis of
> differential methylation experiments

__Author:__ [Nima Hejazi](http://nimahejazi.org)

---

## Description

`methyvim` is an R package that provides facilities for differential methylation
analysis based on _variable importance measures_ (VIMs), a class of statistical
target parameters that arise in causal inference.

The statistical methodology implemented computes targeted minimum loss-based
estimates of well-studied variable importance measures:

1. The average treatment effect (ATE) for discrete exposures/treatments,
2. A nonparametric variable importance measure (NPVI) for continuous exposures
[@chambaz2012estimation].

These methods allow differential methylation effects to be quantified in a
manner that is largely free of assumptions, especially of the variety exploited
in parametric models. __The statistical algorithm consists in several major
steps:__

1. Pre-screening of genomic sites is used to isolate a subset of sites for
which there is evidence of differential methylation. Targeted minimum
loss-based estimates of VIMs are computed for this subset of sites only (for
the sake of computational feasibility). Several screening approaches are
available, adapting core routines from the following R packages:
[`limma`](http://bioconductor.org/packages/release/bioc/html/limma.html),
[`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi),
[`adaptest`](https://github.com/nhejazi/adaptest).
2. Nonparametric VIMs are estimated for the specified parameter, currently using
routines from the R packages
[`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi) to estimate the
NPVI parameter, or [`tmle`](https://CRAN.R-project.org/package=tmle) to
estimate the ATE.
3. Since pre-screening is performed prior to estimating VIMs, we make use of a
multiple testing correction uniquely suited to such settings, specifically we
use the modified marginal Benjamini & Hochberg step-up False Discovery Rate
controlling procedure for multi-stage analyses (FDR-MSA) [@tuglus2009modified].

See @vdl2011targeted for a general description of the framework of targeted
minimum loss-based estimation.

<!--
Note about shrinkage of influence curves, adapting @smyth2004linear.

More exposition here....
-->

---

## Installation

<!--
For standard use, install from [Bioconductor](https://bioconductor.org):
```{r bioc-installation, eval = FALSE}
source("https://bioconductor.org/biocLite.R")
biocLite("methyvim")
```
-->

Install the most recent _stable release_ from GitHub via
[`devtools`](https://www.rstudio.com/products/rpackages/devtools/):

```{r gh-master-installation, eval = FALSE}
devtools::install_github("nhejazi/methyvim")
```

<!--
To contribute, install the _development version_ from GitHub via
[`devtools`](https://www.rstudio.com/products/rpackages/devtools/):

```{r gh-develop-installation, eval = FALSE}
devtools::install_github("nhejazi/methyvim", ref = "develop")
```
-->

---

<!--
## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
## basic example code
```
-->

## Issues

If you encounter any bugs or have any specific feature requests, please [file an
issue](https://github.com/nhejazi/methyvim/issues).

---

## Contributions

It is our hope that `methyvim` will grow to be widely adopted as a tool for the
nonparametric assessment of variable importance in studies of differential
methylation. To that end, contributions are very welcome, though we ask that
interested contributors consult our [`contribution
guidelines`](https://github.com/nhejazi/methyvim/blob/master/CONTRIBUTING.md)
prior to submitting a pull request.

---

## Citation

After using the `methyvim` R package, please cite both of the following:

        @article{hejazi2017methyvim,
          doi = {},
          url = {},
          year  = {2017},
          month = {},
          publisher={Faculty of 1000 Ltd},
          volume = {},
          author = {Hejazi, Nima S. and Hubbard, Alan E. and {van der Laan},
            Mark J.},
          title = {methyvim: Differential methylation analysis with targeted
            nonparametric variable importance measures},
          journal = {F1000Research}
        }

        @misc{hejazi2017methyvim-r,
          doi = {},
          url = {},
          howpublished = {\url{https://github.com/nhejazi/methyvim}},
          year  = {2017},
          author = {Hejazi, Nima S. and Hubbard, Alan E. and {van der Laan},
            Mark J.},
          title = {methyvim: Differential methylation analysis with targeted
            minimum loss-based estimates of variable importance measures}
        }

---

## Related

* [R/`methyvimData`](https://github.com/nhejazi/methyvimData) - R package with
    sample experimental DNA methylation data for use as an example with this
    analysis package.

---

## License

&copy; 2017 [Nima S. Hejazi](http://nimahejazi.org), [Alan E.
Hubbard](http://sph.berkeley.edu/alan-hubbard), [Mark J. van der
Laan](https://www.stat.berkeley.edu/~laan/)

The contents of this repository are distributed under the MIT license. See file
`LICENSE` for details.

---

## References
