---
title: "Targeted Learning of Variable Importance in Differential Methylation"
author: "Nima Hejazi"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document2
bibliography: refs.bib
vignette: >
  %\VignetteIndexEntry{Targeted Learning of Variable Importance in Differential Methylation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

...

---

## Preliminaries

First, we'll load the `methyvim` package and the example data contained in the
`methyvimData` package that accompanies it.

```{r load_pkgs}
if(!("methyvimData" %in% installed.packages())) {
  devtools::install_github("nhejazi/methyvimData")
}
library(methyvimData)
library(methyvim)
```

---

## Quantifying the Effect of an Exposure on DNA Methylation

### The Average Treatment Effect as Variable Importance Measure

...

---

## Quantifying the Effect of DNA Methylation on an Outcome

### A Nonparametric Variable Importance Measure for Continuous Exposures

...[@chambaz2012estimation]...

---

## Session Information

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

---

## References
