---
title: "Targeted Learning of Differential Methylation from an Outcome"
author: "Nima Hejazi"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document2
bibliography: refs.bib
vignette: >
  %\VignetteIndexEntry{Targeted Learning of Differential Methylation from an Outcome}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

...

---

## Preliminaries

First, we'll load the `methyvim` package and the example data contained in the
`methyvimData` package that accompanies it:

```{r load_pkgs}
library(methyvim)
suppressMessages(library(minfi))
suppressMessages(library(SummarizedExperiment))
suppressMessages(library(tmle.npvi))
suppressMessages(library(tmle))
```

```{r prelims}
#options(future.globals.maxSize = 1e12)
options(mc.cores = 2)
set.seed(479253)
```

Now, let's load the data set:

```{r}
library(methyvimData)
data(grsExample)
grsExample
```

---

## Quantifying the Effect of DNA Methylation on an Outcome

### A Nonparametric Variable Importance Measure for Continuous Exposures

@chambaz2012estimation proposed a nonparametric variable importance measure that
is statistically estimatble via targeted minimum loss-based estimation...

```{r methyvim-npvi}
```

...

---

## Session Information

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

---

## References

