---
title: '`methyvim`: an R-package for Variable Importance Measures for
Differential Methylation Analysis'
author:
  - name: Nima S. Hejazi
    affiliation: Address of author1
  - name: Rachael Phillips
    affiliation: orchid.org/0000-0002-8474-591X
  - name: Alan E. Hubbard
    affiliation: Address of author2
  - name: Mark J. van der Laan
    affiliation: Address of author2
abstract:
  We present a general algorithm for the nonparametric estimation of
  treatment/exposure effects on DNA methylation at CpG sites, complete with straightforward statistical inference for such estimates. This approach
  leverages variable importance measures, a class of parameters that arise in
  causal inference and are defined in such a manner that they provide targeted
  estimates of the relative importance of individual CpG site with respect to
  a binary exposure/treatment assigned at the level of subjects. Additionally,
  this procedure is computationally efficient; incorporating a preliminary
  screening step to isolate a subset of sites for which there is cursory
  evidence of differential methylation and a multiple testing correction step
  to control the False Discovery Rate as if all sites were tested. This
  technique for analysis of differentially methylated positions provides an
  avenue to move beyond linear models without the loss of interpretable
  statistical inference.

keywords: DNA methylation, differential methylation, epigenetics, causal
  inference, variable importance, targeted minimum loss-based estimation
bibliography: paper_BiocF1000.bib
output: BiocWorkflowTools::f1000_article
---

# Introduction

DNA methylation is a fundamental epigenetic process known to play an important
role in the regulation of gene expression. DNA CpG methylation involves the
addition of a methyl group ($\text{CH}_3$) to the fifth carbon of the cytosine
ring structure to form 5-methylcytosine. Numerous biological and medical studies
have implicated DNA CpG methylation as playing a role in disease and
development [@robertson2005dna]. Perhaps unsurprisingly then, biotechnologies
have been developed to study the molecular mechanisms of this epigenetic
process. Modern assays, like the Illumina _Infinium_ Methylation assay,
allow for quantitative interrogation of DNA methylation of CpG sites scattered
across the genome at single-nucleotide resolution; moreover, much effort
has been invested, by the bioinformatics community, in the development of tools  
for properly removing technological effects that may contaminate biological
signatures measured by such assays. Despite these advances in both biological
and bioninformatical techniques, most statistical methods available for
differential analysis of data produced by such assays rely on over-simplified
(often generalized linear) models. This is because typical statistical
estimation approaches are not achievable for such high-dimensional data without restrictive modeling assumptions. When these assumptions are violated, estimates
of the population-level effect of an exposure or treatment may have large bias.
Data-adaptive estimation procedures that utilize machine learning can control
for high-dimensional confounders when estimating treatment/exposure effects.
However, interpretable statistical inference (i.e. confidence intervals,
hypothesis tests) resulting from these data-adaptive estimates is challenging [@libbrecht2015machine]. In this paper, we present an alternative such
statistical analysis approaches in the form of nonparametric estimation
procedures that provide simple and well interpretable statistical inference.
Inspired by concepts of counterfactual effects from the causal inference
literature and machine learning algorithms, we provide a computationally
efficient technique for obtaining targeted estimates of nonparametric _variable importance measures_ (__VIM__) for a pre-screened set of CpGs, controlling for
the False Discovery Rate (FDR) as if all sites were tested. Under counterfactual assumptions, these estimators have asymptotic normal sampling distributions
and thus provide reliable inference. Counterfactuals are "counter"-fact, defined
in terms of an unrealistic yet ideal experiment. In the context of DNA
methylation and for a binary exposure/treatment, we define the counterfactual
outcomes as the Beta-value/M-value a CpG site would have if all subjects were
exposed/treated and the Beta-value/M-value a CpG site would have if all
subjects were not exposed/treated. Of course, these counterfactual outcomes are
impossible to observe but they are very helpful for us to asses the importance
of a CpG site relative to a subjects treatment/exposure and they can be
statistically estimated [@van2006statistical]. Our VIM definition uses the
concepts of counterfactuals and the framework for estimation is built on
targeted maximum likelihood estimation or targeted minimum loss-based estimation
(TMLE) [@van2006targeted]. This methodology targets the individual importance
of each CpG site by taking advantage of the prediction power of machine learning
algorithms. The mathematical properties of this approach have been explored in [@tuglus2008targeted] and we focus on our contribution to it's application,
`methyvim`, an R package that accommodates this technique for differential
methylation analysis.


For a general discussion of the framework of targeted minimum loss-based
estimation and the role this approach plays in statistical causal inference, the
interested reader is invited to consult @vdl2011targeted and @vdl2017targeted.
For a more general introduction to (statistical) causal inference,
@pearl2009causality and @hernan2018causal may be of interest.

# Methods

## Implementation

The core functionality of this package is made available via the eponymous
`methyvim` function, which implements a statistical algorithm designed to
compute targeted estimates of VIMs, defined in such a way that the VIMs
represent parameters of scientific interest in computational biology
experiments; moreover, these VIMs are defined such that they may be estimated in
a manner that is very nearly assumption-free, that is, within a _fully
nonparametric statistical model_. __The statistical algorithm consists in
several major steps:__

1. Pre-screening of genomic sites is used to isolate a subset of sites for
  which there is cursory evidence of differential methylation. For the sake of
  computational feasibility, targeted minimum loss-based estimates of VIMs are
  computed only for this subset of sites. Currently, the available screening
  approach adapts core routines from the
  [`limma`](http://bioconductor.org/packages/release/bioc/html/limma.html) R
  package. Future releases will support functionality from other
  packages (e.g., [`randomForest`](https://cran.r-project.org/web/packages/randomForest/index.html),
  [`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi)). Following the
  style of the function for performing screening via `limma`, users may write
  their own screening functions and are invited to contribute such functions to
  the core software package by opening pull requests at the GitHub repository.

2. Nonparametric estimates of VIMs, for the specified parameter of interest, are
  computed at each of the CpG sites passing the screening step. The VIMs are
  defined in such a way that the estimated effects is of an exposure/treatment
  on the methylation status of a target CpG site, controlling for the observed
  methylation status of the neighbors of that site. Currently, routines are
  adapted from the [`tmle`](https://CRAN.R-project.org/package=tmle) R package.
  Future releases will support doubly-robust estimates of these VIMs (via the
  [`drtmle`](https://cran.r-project.org/web/packages/drtmle/index.html) package)
  and add parameters for continuous treatments/exposures (via the
  [`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi) package).

3. Since pre-screening is performed prior to estimating VIMs, we make use of a
  multiple testing correction uniquely suited to such settings. Due to the
  multiple testing nature of the estimation problem, a variant of the Benjamini
  & Hochberg procedure for controlling the False Discovery Rate (FDR) is applied
  [@benjamini1995controlling]. Specifically, we apply the "modified marginal
  Benjamini & Hochberg step-up False Discovery Rate controlling procedure for
  multi-stage analyses" (FDR-MSA), which is guaranteed to control the FDR as if
  all sites were tested (i.e., without screening) [@tuglus2009modified].

#### Parameters of Interest

For the CpG sites that passed the pre-screening step, the user chooses the
parameter of interest that will be computed at each site. In all cases, an
estimator of the parameter of interest is constructed via targeted minimum
loss-based estimation.

For discrete-valued treatments or exposures:

* The _average treatment effect_ (ATE): The effect of a binary exposure or
  treatment on the observed methylation at a target CpG site is estimated,
  controlling for the observed methylation at all other CpG sites in the same
  neighborhood as the target site, based on an additive form. Often denoted
  $\psi_0 = \psi_0(1) - \psi_0(0)$, the parameter estimate represents the
  __additive difference__ in methylation that would have been observed at the
  target site had all observations received the treatment versus the
  counterfactual under which none received the treatment.

* The _relative risk_ (RR): The effect of a binary exposure or treatment on the
  observed methylation at a target CpG site is estimated, controlling for the
  observed methylation at all other CpG sites in the same neighborhood as the
  target site, based on a geometric form. Often denoted,
  $\psi_0 = \frac{\psi_0(1)}{\psi_0(0)}$, the parameter estimate represents the
  __multiplicative difference__ in methylation that would have
  been observed at the target site had all observations received the treatment
  versus the counterfactual under which none received the treatment.

Support for continuous-valued treatments or exposures is _planned but not yet
available_, though work is underway to incorporate into our methodology the
following

* A _nonparametric variable importance measure_ (NPVI) [@chambaz2012estimation]:
  The effect of continuous-valued exposure or treatment (the observed
  methylation at a target CpG site) on an outcome of interest is estimated,
  controlling for the observed methylation at all other CpG sites in the same neighborhood as the target (treatment) site, based on a parameter that
  compares values of the treatment against a reference value taken to be the
  null. In particular, the implementation provided is designed to assess the
  effect of differential methylation at the target CpG site on a (typically) phenotype-level outcome of interest (e.g., survival), in effect providing an nonparametric evaluation of the impact of methylation at the target site on
  said outcome.

We have adopted a class `methytmle` to help organize the functionality within
this package. The `methytmle` class builds upon the `GenomicRatioSet` class
provided by the `minfi` package so all of the slots of `GenomicRatioSet` are
contained in a `methytmle` object. The new class introduced in the `methyvim`
package includes several new slots:

* `call` - the form of the original call to the `methyvim` function.
* `screen_ind` - indices identifying CpG sites that pass the screening process.
* `clusters` - non-unique IDs corresponding to the manner in wich sites are
  treated as neighbors. These are assigned by genomic distance (bp) and respect
  chromosome boundaries (produced via a call to `bumphunter::clusterMaker`).
* `var_int` - the treatment/exposure status for each subject. Currently, these
  must be binary, due to the definition of the supported targeted parameters.
* `param` - the name of the target parameter from which the estimated VIMs are
  defined.
* `vim` - a table of statistical results obtained from estimating VIMs for each
  of the CpG sites that pass the screening procedure.
* `ic` - the measured array values for each of the CpG sites passing the
  screening, transformed into influence curve space based on the chosen target
  parameter.

The interested analyst might consider consulting the documentation of the
`minfi` package for an in-depth description of all of the other slots that
appear in this class [@aryee2014minfi].

We refer the reader to the package vignette, "`methyvim`: Variable Importance
Measures for Differential Methylation" hosted with the package on Bioconductor,
for a detailed description of the methods used in this package.
(NEED CITATION FOR VIGNETTE)

## Operation

A standard computer with the latest version of R and Bioconductor 3.6 installed
will handle applications of the `methyvim` package.

# Use Cases

To examine the practical applications and the full set of utilities of the
`methyvim` package, we will use a publicly available example data set produced
by the Illumina 450K array, from the `minfiData` R package.

### Preliminaries: Setting up the Data

We begin by loading the package and the data set. After loading the data, which
comes in the form of a raw `MethylSet` object, we perform some further
processing by mapping to the genome (with `mapToGenome`) and converting the
values from the methylated and unmethylated channels to Beta-values
(via `ratioConvert`). These two steps together produce an object of class `GenomicRatioSet`, provided by the `minfi` package.

```{r setup-minfidata}
suppressMessages(library(minfiData))
data(MsetEx)
mset <- mapToGenome(MsetEx)
grs <- ratioConvert(mset)
grs
```
We can create an object of class `methytmle` from any `GenomicRatioSet` object
simply invoking the S4 class constructor:

```{r make-methytmle}
mtmle <- .methytmle(grs)
```
Additionally, a `GenomicRatioSet` can be created from a matrix with the function `makeGenomicRatioSetFromMatrix` provided by the `minfi` package.

### Differential Methylation Analysis

For this example analysis, we'll treat the condition of the patients as the
exposure/treatment variable of interest. The `methyvim` function requires that
this variable either be `numeric` or easily coercible to `numeric`. To
facilitate this, we'll simply convert the covariate (currently a `character`):

```{r, minfidata-maketx}
var_int <- (as.numeric(as.factor(colData(grs)$status)) - 1)
```
__n.b.__, the re-coding process results in "normal" patients being assigned a
value of 1 and cancer patients a 0.

Now, we are ready to analyze the effects of cancer status on DNA methylation
using this data set. We proceed as follows with a targeted minimum loss-based
estimate of the Average Treatment Effect.

```{r, minfidata-methyvim}
suppressMessages(
  methyvim_cancer_ate <- methyvim(data_grs = grs, var_int = var_int,
                                  vim = "ate", type = "Beta", filter = "limma",
                                  filter_cutoff = 0.20, obs_per_covar = 2,
                                  parallel = FALSE, sites_comp = 125,
                                  tmle_type = "glm"
                                 )
)
```
Note that we set the `obs_per_covar` argument to a relatively low value (2,
where the recommended default is 20) for the purposes of this example as the
sample size is only 10. We do this only to exemplify the estimation procedure
and it is important to point out that such low values for `obs_per_covar` will
compromise the quality of inference obtained because this setting directly
affects the definition of the target parameter.

Further, note that here we apply the `glm` flavor of the `tmle_type` argument,
which produces faster results by fitting models for the propensity score and
outcome regressions using a limited number of parametric models. By contrast,
the `sl` (for "Super Learning") flavor fits these two regressions using highly
nonparametric and data-adaptive procedures (i.e., via machine learning).
Obtaining the estimates via GLMs results in each of the regression steps
being less robust than if nonparametric regressions were used.

We can view a table of results by examining the `vim` slot of the produced
`methytmle` object.

```{r vim-cancer-ate}
head(slot(methyvim_cancer_ate, "vim"))
```

Finally, we may compute FDR-corrected p-values, by applying a modified procedure
for controlling the False Discovery Rate for multi-stage analyses (FDR-MSA)
[@tuglus2009modified]. We do this by simply applying the `fdr_msa` function.

```{r fdr-msa}
fdr_p <- fdr_msa(pvals = slot(methyvim_cancer_ate, "vim")$pval,
                 total_obs = nrow(methyvim_cancer_ate))
```

Having explored the results of our analysis numerically, we now proceed to use
the visualization tools provided with the `methyvim` R package to further
enhance our understanding of the results.

### Visualization of Results

While making allowance for users to explore the full set of results produced by
the estimation procedure (by way of exposing these directly to the user), the
`methyvim` package also provides _three_ (3) visualization utilities that
produce plots commonly used in examining the results of differential methylation
analyses.

A simple call to `plot` produces side-by-side histograms of the raw p-values
computed as part of the estimation process and the corrected p-values obtained
from using the FDR-MSA procedure.

```{r methyvim-pvals-both}
plot(methyvim_cancer_ate)
```
__Remark:__ The plots displayed above may also be generated separately by
explicitly setting the argument "type" to `plot.methytmle`. For a plot of the
raw p-values, specify `type = "raw_pvals"`, and for a plot of the FDR-corrected
p-values, specify `type = "fdr_pvals"`.

While histograms of the p-values may be generally useful in inspecting the
results of the estimation procedure, a more common plot used in examining the
results of differential methylation procedures is the volcano plot, which plots
the parameter estimate along the x-axis and $-\text{log}_{10}(\text{p-value})$
along the y-axis. We implement such a plot in the `methyvolc` function:

```{r methyvim-volcano}
methyvolc(methyvim_cancer_ate)
```

The purpose of such a plot is to ensure that very low (possibly statistically
significant) p-values do not arise from cases of low variance. This appears to
be the case in the plot above (notice that most parameter estimates are _near
zero_, even in cases where the raw p-values are quite low).

Yet another popular plot for visualizing effects in such settings is the
heatmap, which plots estimates of the raw methylation effects (as measured by
the assay) across subjects using a heat gradient. We implement this in the
`methyheat` function:

```{r methyvim-heatmap}
methyheat(methyvim_cancer_ate)
```

Invoking `methyheat` in this manner produces a plot of the top sites ($25$, by
default) based on the raw p-value, using the raw methylation measures in the
plot. This uses the exceptional `superheat` R package [@barter2017superheat].

# Summary <!-- Optional - only if NO new datasets are included -->

This section is required if the paper does not include novel data or analyses.
It allows authors to briefly summarize the key points from the article.


# Software availability
This section will be generated by the Editorial Office before publication.
Authors are asked to provide some initial information to assist the Editorial
Office, as detailed below.

1. URL link to where the software can be downloaded from or used by a non-coder (AUTHOR TO PROVIDE; optional)
2. URL link to the author's version control system repository containing the source code (AUTHOR TO PROVIDE; required)
3. Link to source code as at time of publication (*F1000Research* TO GENERATE)
4. Link to archived source code as at time of publication (*F1000Research* TO GENERATE)
5. Software license (AUTHOR TO PROVIDE; required)

Software available from:

Latest source code: https://github.com/nhejazi/methyvim

Archived source code as at time of publication:

Software license:


# Author contributions

NH designed and implemented the tool, applied the tool to the use case, and
drafted the vignette. RP helped design the tool and drafted the article.

# Competing interests
All financial, personal, or professional competing interests for any of the
authors that could be construed to unduly influence the content of the article
must be disclosed and will be displayed alongside the article. If there are no
relevant competing interests to declare, please add the following: 'No competing
interests were disclosed'.

# Grant information
Please state who funded the work discussed in this article, whether it is your
employer, a grant funder etc. Please do not list funding that you have that is
not relevant to this specific piece of research. For each funder, please state
the funder’s name, the grant number where applicable, and the individual to whom
the grant was assigned. If your work was not funded by any grants, please
include the line: 'The author(s) declared that no grants were involved in
supporting this work.'

# Acknowledgments
This section should acknowledge anyone who contributed to the research or the
article but who does not qualify as an author based on the criteria provided
earlier (e.g. someone or an organization that provided writing assistance).
Please state how they contributed; authors should obtain permission to
acknowledge from all those mentioned in the Acknowledgments section.

Please do not list grant funding in this section.

# References
